### 2014-10-14 ApTranscriptome_TR.Rmd
cleaning up files by removing analyses no longer referred to in ms

- compare profiles of the temperature of maximum expression for thermally-responsive transcripts
- compare basal expression at optimal temperature for thermally-responsive genes
- compare thermal stability of *Intermediate* genes
- compare thermal sensitivity of *Bimodal* genes
- compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
- evaluate the extent to which thermal reaction norms differ between species by mean and shape

@@

### Compare thermal stability of *Intermediate* genes

*Intermediate* genes are those that have expression that is shut-down as conditions become stressful, likely non-essential molecular processes. We hypothesized that if the more southern *Ar* species was more thermally-tolerant than *A22*, transcripts with *Intermediate* expression (10-30C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each species.

```{r thermal_breadth}
# extract 'Intermediate' expressed transcripts for A22
A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
length(A22.int.lm)

# apply `transcriptSD` function to all transcripts
A22.int.sd <- unlist(Map(transcriptSD, A22.int.lm, colony = "A22"))
A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)

# repeat for Ar
Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))
Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)

# T-test comparing standard deviation of expression between colonies
(t.varint <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
```

Consistent with our hypothesis, *Intermediate* transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.

```{r plot_thermal_breadth, echo=FALSE, eval=TRUE}
# prep data for plotting
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

# plot thermal breadth
g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

### Compare thermal sensitivity of *Bimodal* genes

As the converse of the above hypothesis, a species that is especially thermally-sensitive is likely to activate expression of molecular processes for thermal tolerance more quickly. We tested this using the same approach as for the *Intermediate* transcripts, but using the inverse of the *Bimodal* expressed transcripts.

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 species
A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
length(A22.bim.lm)

# apply `transcriptSD` function to all transcripts
A22.bim.sd <- unlist(Map(transcriptSD, A22.bim.lm, colony = "A22"))

# repeat for Ar 
Ar.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]]
length(Ar.bim.lm)
Ar.bim.sd <- unlist(Map(transcriptSD, Ar.bim.lm, colony = "Ar"))

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
(t.varbim <- t.test(Ar.bim.sd, A22.bim.sd))
```

No difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.

### Compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes

@@

Repeat analysis for when **Intermediate** genes are turned *off*.

@@

# grid.arrange
grid.arrange(gg2, gg3, nrow = 2)
```

@@

### Evaluate the extent to which differences in thermal reaction norms are due to mean or shape

For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect? To test this, I rougly follow `r citet("10.1086/675302")` by defining differences among reation norms for individual genes due to changes in the overall mean, slope, curvature and all higher-order shape differences (i.e. wiggle).

- *Mean, M*: overall difference in the mean expression value across all temperatures
- *Slope, S*: difference in overall slope
- *Curvature, C*: average difference in curvature of the reaction norm
- *Wiggle, W*: variability in shape not captured by the previous three measures

```{r varshape, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)

# delta mean
boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))

t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean), paired = TRUE)
t.mean

# delta slope
boxplot(log(varshape.out$A22.slope), log(varshape.out$Ar.slope))
t.slope <- t.test(log(varshape.out$A22.slope + 0.1), log(varshape.out$Ar.slope + 0.1), paired = TRUE)
t.slope

# delta curvature
boxplot(log(varshape.out$A22.curve), log(varshape.out$Ar.curve))
t.curvature <- t.test(log(varshape.out$A22.curve + 0.1), log(varshape.out$Ar.curve + 0.1), paired = TRUE)
t.curvature

# delta wiggle - small values so use un-transformed
boxplot(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
t.wiggle <- t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle, paired = TRUE)
t.wiggle
```

Reaction norms differ between species by mean and slope, but not by curvature or wiggle.

Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means and slope, as curvature and wiggle did not differ between species.

```{r partition_reaction_norm_variation}
# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
# mean = mean1 - mean2
varshape.out$lmean <- log(varshape.out$A22.mean) - log(varshape.out$Ar.mean)

# slope = slope1 - slope2
varshape.out$lslope <- log(varshape.out$A22.slope + 0.1) - log(varshape.out$Ar.slope + 0.1)

# take absolute value of each value and sum to get total differences between reaction norms
varshape.out$ltotal <- abs(varshape.out$lmean) + abs(varshape.out$lslope)

# variation in reaction norms due to differences in mean
varshape.out$prop.lmean <- abs(varshape.out$lmean) / varshape.out$ltotal
varshape.out$prop.lslope <- abs(varshape.out$lslope) / varshape.out$ltotal

# Mean proportion and 95% CI of total variation of each measure
mean(varshape.out$prop.lmean)
(qlmean <- quantile(varshape.out$prop.lmean, probs = c(0.05, 0.5, 0.95)))

mean(varshape.out$prop.lslope)
(qlslope <- quantile(varshape.out$prop.lslope, probs = c(0.05, 0.5, 0.95)))
```

From this analysis, about 80% of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.

@@

```{r rData, echo=FALSE}
# save RData for manuscript file
save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
```

>>>>

- compare profiles of the temperature of maximum expression for thermally-responsive transcripts
- compare basal expression at optimal temperature for thermally-responsive genes
- compare degree of uregulation to constitutive expression
- compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
- compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures

@@

### Comparison of the degree of upregulation against constitutive expression

To test the genetic assimilation hypothesis, I compare the degree of upregulation of a transcript against constitutive expression at optimal temperatures.

```{r induction_vs_constitutive}
Ap.constitutive <- resp.TPM.dt.sub.pred[resp.TPM.dt.sub.pred$val == ]

### Compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures

@@

### Compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures

This analysis is comparable to the one above, but for the *Intermediate* genes that are downregulated, split into the low and high temperature extremes.

@@

# grid.arrange
#tiff("results/crit_temp_regulation_Fig2.tiff", compression = "lzw", bg = "white")
grid.arrange(gg2, gg3, nrow = 2)
#dev.off()
```

@@

@@

```{r rData, echo=FALSE}
# save RData for manuscript file
save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.T_off, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
```
